 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-04-23 paper.Rmd
update case study one

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

@@

In the following section, we provide a brief overview of the goals
of DR methods with respect to the data analyst, in addition to their high level the 
mathematical details. To begin we suppose the data is in the form 
rectangular matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, 
where $n$ is the number of observations in $p$ dimensions. 
The purpose of any DR algorithm is to find low-dimensonal
representation of the data, $Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that
$Y$ is an $n \times d$ matrix where $d \ll p$. The hope of the analyst is that
the DR procedure to produces $Y$ will remove noise in the original dataset while
retaining latent structure.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input 
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation. 
Here we restrict our attention to two recent methods that are commonly
used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pairwise distances between   $\mathbf{x_i}$ and 
$\mathbf{x_j}$  to similiarties using a Gaussian kernel:

@@

The conditional probabilities are then normalised and symmetrised to form a 
joint probability  distrubition via averaging:

@@

Next, the target low-dimensional space, $Y$, pairwise distances between 
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability 
distrubtion using a Cauchy kernel:

@@

The resulting configuration $Y$ is
the one that minimizes the Kullback-Leibler divergence between 
the probability distrubtions formed via similiarties of observations in $X$, $\mathcal{P}$ 
and similiarities of observations in $Y$, $\mathcal{Q}$:

@@

So when the the loss function is minimised this corresponds to large
attractive forces, that is, the pairwise distances in $Y$ are small when
there are non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The
repulsive force should also be small when the loss function is minimised, 
that is, when pairwise distances in $Y$ are large irregardless of the magnitude
of the corresponding distances in $X$.

Taken together, these details reveal the sheer number of decisions that an 
anlayst must make. How does one choose the perplexity? How should the parameters
that control the optimisation of the loss function (done with stochastic
gradient descent), like the number of iterations, or early exaggeration (
a multiplier of the attractive force at the beginning of the optimisation),
or the learning rate be selected?  It is a known
problem that t-SNE can have trouble recovering topology and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is
cluster orientation a recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. However,
the cluster sizes and their relative orientation from a $t-SNE$ view can be 
misleading perceptually, due to the algorithms emphasis on locality.

Another recent method UMAP has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. It appears to suffer from the same perceptual issues
as t-SNE, however it supposedly preservers global structure better
than t-SNE [@Coenen2019-to].

@@

The next section steps through three use cases of our approach 
using simulated and real data. In the first case study, we step through
data that lies on tree structure (similar to trajectories in single
cell transcriptomics) to that highlights an iterative approach to modifying
t-SNE views based on what we learn in the high-dimensional space. We show
how through simple linked brushing, we can see when the assumptions
of continuity of the embedding map are broken. In the second case study,
we explore a well-clustered single cell transcriptomics data of mice retina...
Finally, in the last case study we explore a subset of the QuickDraw dataset,
we use `liminal` to identify instances of images from different labels that
appear similiar.

@@

Moreover,
we run t-SNE using the default arguments on the complete data (this keeps the first 50 PCs, sets the perpelxity to equal 30 and performs random initalisation) using the `Rtsne` package [@Rtsne]. We then create a linked tour with t-SNE layout with `liminal` as shown in the following video :

<embededded video 1>

From the video, we see that the t-SNE view has been unable to recapitulate the topology of the tree - the backbone (blue) branch has been split into three fragments. We can see this immediately via the linked highlighting over both plots. If we click on the legend for the zeroth branch, the blue colored points on each view are higlighted and the remaining points are made transparent. From, here it becomes apparent from the tour view that the blue branch forms the backbone of the tree and is connected to all other branches.
We can also again see that cluster sizes formed via t-SNE can be misleading; from the tour view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result.

@@

The video shows that this selection of parameters results
in the the tips of the branch

>>>>

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to colour points, and
highlights nearest neighbours [@Smilkov2016-hp].

@@

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find low-dimensional representation of the data, $Y = [\mathbf{y_1},
\dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$ matrix where $d \ll p$.
The hope of the analyst is that the DR procedure to produces $Y$ will remove
noise in the original dataset while retaining latent structure.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Here we restrict our attention to two recent methods that are commonly used in
the literature: t-distributed stochastic neighbour embedding (t-SNE) and
uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
the pair-wise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

@@

The conditional probabilities are then normalised and symmetrised to form a
joint probability distribution via averaging:

@@

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

The resulting configuration $Y$ is the one that minimizes the Kullback-Leibler
divergence between the probability distributions formed via similarities of
observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
$\mathcal{Q}$:

@@

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pair-wise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pair-wise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

Taken together, these details reveal the sheer number of decisions that an
analyst must make. How does one choose the perplexity? How should the
parameters that control the optimisation of the loss function (done with
stochastic gradient descent), like the number of iterations, or early
exaggeration ( a multiplier of the attractive force at the beginning of the
optimisation), or the learning rate be selected?  It is a known problem that
t-SNE can have trouble recovering topology and that configurations can be
highly dependent on how the algorithm is initialised and parameterized
[@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is cluster
orientation a recent theoretical contribution by @Linderman2019-dq proved that
t-SNE can recover spherical and well separated cluster shapes, and proposed new
approaches for tuning the optimisation parameters. However, the cluster sizes
and their relative orientation from a $t-SNE$ view can be misleading
perceptually, due to the algorithms emphasis on locality.

Another recent method UMAP has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. is a method that is related to
LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
Its main differences are that it uses a different cost function (cross entropy)
which is optimized using stochastic gradient descent and defines a different
kernel for similarities in the low dimensional space. Due to it's computational
speed it's possible to generate UMAP embeddings in more than three dimensions.
It appears to suffer from the same perceptual issues as t-SNE, however it
supposedly preservers global structure better than t-SNE [@Coenen2019-to].

@@

The next section steps through three use cases of our approach using simulated
and real data. In the first case study, we step through data that lies on tree
structure (similar to trajectories in single cell transcriptomics) to that
highlights an iterative approach to modifying t-SNE views based on what we
learn in the high-dimensional space. We show how through simple linked
brushing, we can see when the assumptions of continuity of the embedding map
are broken. In the second case study, we explore a well-clustered single cell
transcriptomics data of mice retina... Finally, in the last case study we
explore a subset of the QuickDraw dataset, we use `liminal` to identify
instances of images from different labels that appear similar.

@@

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation) using the `Rtsne` package [@Rtsne]. We then create a linked
tour with t-SNE layout with `liminal` as shown in the following video :

<embedded video 1>

From the video, we see that the t-SNE view has been unable to recapitulate the
topology of the tree - the backbone (blue) branch has been split into three
fragments. We can see this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. We can also
again see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result.

@@

The video shows that this selection of parameters results in the the tips of
the branches (the three black dots in figure \@ref(fig:fake-trees)) being split
into three clusters representing the final branches of the tree. However, there
are perceptual issues following the placement of the three groupings on the
t-SNE view that become apparent via simple linked brushing. If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree.
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity.
